---
title: "Building the gazetteer snapshot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building the gazetteer snapshot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(georefapp)
```

While you georeference a locality, `georefapp` shows you places elsewhere whose
name resembles it. Those come from the **gazetteer**: a local, read-only
snapshot of the locality dictionary, built once and opened as a file. The app
never queries the live database while you work.

# Quick start

## Build a snapshot

Credentials are read from the environment, never written into a script. Put
them in `~/.Renviron` (or the project's own `.Renviron`, which is
gitignored) and restart R:

```
MYDB_USER=your_user
MYDB_PASS=your_password
```

```{r build, eval = FALSE}
con <- DBI::dbConnect(
  RPostgres::Postgres(),
  host = "dg474899-001.dbaas.ovh.net", port = 35699L,
  dbname = "plots_transects",
  user = Sys.getenv("MYDB_USER"),
  password = Sys.getenv("MYDB_PASS")
)

gazetteer_snapshot_build(con, "rainbio-gazetteer.sqlite", overwrite = TRUE)
```

About half a minute for the full RAINBIO dictionary, giving a 126 MB file.

`file` follows the base R convention and names the file to write, as in
`saveRDS()`. A directory is accepted too — the default name is written inside
it, and the resolved name is returned invisibly either way.

`src` takes a data frame just as readily as a connection, so a dictionary that
does not live in this database needs no different call — only the same columns.

## Use it

```{r use, eval = FALSE}
launch(gazetteer = "rainbio-gazetteer.sqlite")

# or set it for a session without launching
use_gazetteer("rainbio-gazetteer.sqlite")
```

Running with no gazetteer is fully supported, and is the normal state for most
of Central Africa — the app shows "no similar locality found", which is what a
real dictionary returns for a place it does not know.

## Rebuild it when the source changes

After new records are loaded, synonymy is curated, or the view definition is
edited. The view behind the dictionary is materialized, so refresh it first —
otherwise the snapshot faithfully reproduces a stale view:

```{r refresh, eval = FALSE}
DBI::dbExecute(con, "REFRESH MATERIALIZED VIEW rainbio_gazetteer_localities")
gazetteer_snapshot_build(con, "rainbio-gazetteer.sqlite", overwrite = TRUE)
```

Each rebuild gets a new snapshot identifier, stamped onto every decision made
while it is loaded — so a decision always names the dictionary state it was
made against.

# How it works

## Why a local file

The dictionary lives in PostgreSQL on a server in another country, and the whole
of it is copied into SQLite rather than queried. Three reasons, in order of how
much they matter:

1. **A georeferencing session must not stall.** You decide a few hundred
   localities in a sitting, one lookup each. A slow or unreachable database
   would interrupt the loop at the worst possible moment.
2. **A decision has to be reconstructible.** If the dictionary is a moving
   target on a server, "what was on screen" is unanswerable a year later. If it
   is a file, it is a file.
3. **The search needs an index the source does not have** — an inverted index
   over the whole corpus, built once at snapshot time rather than per query.

The source is one row per locality, keyed on the *canonical* id, so several
spellings can collapse into one entry. Two things about it are easy to misread:

- **The coordinate is a median, not a georeference.** Nobody decided it; it is
  the middle of whatever the collectors wrote down. This is why candidates are
  plotted as bare markers with no uncertainty circle — a radius would dress a
  summary statistic up as a measurement.
- **Localities with no coordinates are kept.** Of 360,583 entries, 113,185 have
  none. A similar name that is itself ungeoreferenced is still evidence, usually
  that the same place is filed under another spelling. The app lists them greyed
  and does not plot them.

## Step 1 — tokenising

Each locality string is reduced to a set of words.

```{r tokens}
locality_tokens("P.N. Odzala. Mbandza, village sur la Lékoli, au NNW de Mbomo")
```

Characters are normalised — transliterated to ASCII, lowercased, punctuation
turned into spaces — so `Lékoli` and `Lekoli` become one token. Tokens are
deduplicated within a locality: a locality is a name, not a document, so a word
appearing twice is not stronger evidence than a word appearing once.

Two kinds of word are dropped. Single characters go, because the bearings they
usually stand for are in so many localities that their weight would be
negligible anyway. Hedges go too:

```{r hedges}
locality_tokens("env. de Kribi")
locality_tokens("30 km env. ESE of Kribi")
```

The hedge list is the one `normalise_locality()` already strips from the front
of a key, applied wherever the word occurs. The justification is the definition
already in use: a word that does not change *which place is meant*. This is not
cosmetic — `env.` is rare enough across the corpus that the weighting below
treated it as strong evidence, and `env. de Kribi` used to match **`Env. de
Léopoldville`** at 50%.

Displacement words are deliberately *not* dropped. `km`, `north`, `ESE` change
the place, so they stay.

## Step 2 — weighting by rarity

A shared word is worth something in proportion to how surprising it is.
`village` is in tens of thousands of localities and says almost nothing;
`Odzala` is in a few dozen and says almost everything. The measure is inverse
document frequency — for a token in `df` localities out of `N`:

$$\mathrm{idf} = \log\left(1 + \frac{N}{df}\right)$$

```{r idf}
idf <- function(df, N = 360582) log1p(N / df)
data.frame(
  token = c("de", "village", "riviere", "kribi", "odzala", "mbandza"),
  df    = c(180000,   15000,     4000,     900,      40,        12),
  idf   = round(idf(c(180000, 15000, 4000, 900, 40, 12)), 2)
)
```

A word appearing in exactly one locality scores 12.8 — four times `village`,
eighteen times `de`. That separation is what makes a stoplist unnecessary, and
it has to be unnecessary here: the corpus mixes French, English, Portuguese and
local languages, and no hand-written list would cover all four. The corpus
decides instead.

Words in more than 5% of the corpus are counted but not indexed: they cannot
discriminate, and searching them would mean scanning tens of thousands of
entries for nothing. On the real corpus that prunes 10 tokens out of 133,198 —
words like `de` and `of`. Everything else, `village` included, stays indexed and
is handled by its weight rather than by exclusion.

## Step 3 — scoring a query

The builder takes a data frame as readily as a connection, so this runs on eight
real Mbandza localities instead of 360,000:

```{r demo}
demo <- data.frame(
  locality_id = 1:8,
  locality_verbatim = c(
    "P.N. Odzala. Mbandza, village sur la Lekoli, au NNW de Mbomo",
    "Parc National d'Odzala. Village Mbandza.",
    "parc national d'odzola, ancien champ du cacao, village mbandza",
    "Congo Brazzaville. Layon Mbandza (km I)",
    "Yangambi, village de la riviere",
    "Parc National de Lope, village Ipassa",
    "Kribi, village de peche",
    "Route de Ouesso, village Liouesso"),
  is_georeferenced  = c(1L, 1L, 0L, 0L, 1L, 1L, 1L, 0L),
  decimal_latitude  = c(0.62, 0.63, NA, NA, 0.81, -0.20, 2.94, NA),
  decimal_longitude = c(14.81, 14.80, NA, NA, 24.49, 11.60, 9.91, NA),
  n_records         = c(12L, 5L, 3L, 1L, 40L, 7L, 9L, 2L),
  n_records_georef  = c(12L, 5L, 0L, 0L, 40L, 7L, 9L, 0L),
  n_distinct_coords = c(4L, 2L, NA, NA, 11L, 3L, 5L, NA),
  span_deg          = c(0.05, 0.01, NA, NA, 0.4, 0.02, 0.1, NA),
  n_name_variants   = 1L
)
demo$locality_key <- normalise_locality(demo$locality_verbatim)

f <- file.path(tempdir(), "demo-gazetteer.sqlite")
unlink(f)
gazetteer_snapshot_build(demo, f, snapshot_id = "demo-1", df_max_frac = 0.5)

gaz <- gazetteer_open(f)
hits <- gazetteer_search(gaz, "Parc National d'Odzala, village Mbandza",
                         min_coverage = 0.3)
hits[, c("locality_verbatim", "is_georeferenced", "coverage", "overlap")]
```

Every locality sharing an indexed token with the query is scored on the summed
weight of the tokens they share, and two figures come out:

- **coverage** — the share of the *query's* weight the candidate accounts for.
  The primary ranking, because the question is "where are the localities that
  mention this place", not "which localities are worded like this one".
- **overlap** — the weighted Jaccard index, which penalises a candidate for
  saying more than the query did. It breaks ties.

The georeferenced entry comes first; its ungeoreferenced near-twin, spelled
`odzola` and carrying no coordinates, is right behind it. That is exactly the
case the panel exists to surface.

Overlap penalises *information*, not length — worth knowing before you read a
ranking:

```{r overlap}
gazetteer_search(gaz, "Mbandza", min_coverage = 0.3)[
  , c("locality_verbatim", "coverage", "overlap")]
```

All four contain the whole query, so overlap decides. `Congo Brazzaville. Layon
Mbandza (km I)` is by some way the shortest and does not come first: every word
in it bar `mbandza` is unique in this corpus and carries maximum weight, making
it a specific other place rather than a variant of the query.

## What it cannot do

Matching is exact on words, so a misspelling is its own rare token:

```{r typo}
gazetteer_search(gaz, "odzola", min_coverage = 0.3)$locality_verbatim
```

One hit — the entry sharing the typo. The correctly spelled Odzala localities
are invisible to it, and vice versa. Closing that gap needs a character-level
measure over the token vocabulary; token rarity alone cannot repair spelling.

A query made only of common words is reported as unsearchable rather than
matched against everything:

```{r common}
none <- gazetteer_search(gaz, "village")
nrow(none)
attr(none, "query")$too_common
```

```{r cleanup, include = FALSE}
gazetteer_close(gaz)
unlink(f)
```

---

A longer walkthrough — every step of the build, the schema, benchmark numbers —
is kept with the project's internal notes rather than shipped here, and
`R/gazetteer.R` and `R/tokens.R` carry the same reasoning in comments.
